---
title: "Families and model types in fbrglm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Families and model types in fbrglm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>",
    fig.width  = 6,
    fig.height = 4
)
set.seed(20260201)
```

## Overview

`fbrglm` exposes a single formula/data interface for regularized GLMs
and reuses one set of S3 methods across model types. The common
operations are:

- `fbrglm(formula, data, family, lambda, ...)`
- `print()` / `summary()` / `coef()` / `nobs()`
- `predict(newdata, type = c("link", "response", "class"), newoffset)`
- `plot()` (delegates to the underlying `glmnet` / `cv.glmnet` plot)
- `as_glmnet()` / `as_cv_glmnet()`

Classical standard errors, *z* values, and *p* values are **not**
produced under `infer = "none"` (the only inference mode currently
implemented). The package returns regularized point estimates;
inference modes are the next milestone.

```{r setup}
library(fbrglm)
```

## Summary of supported model types

| Model                                   | fbrglm formulation                                                                   | Response               | Status               | Note                                                                                                  |
|-----------------------------------------|--------------------------------------------------------------------------------------|------------------------|----------------------|-------------------------------------------------------------------------------------------------------|
| Linear regression                       | `family = "gaussian"`                                                                | continuous `y`         | **supported**        | standard Gaussian model                                                                               |
| Logistic regression                     | `family = "binomial"`                                                                | 0 / 1 `y`              | **supported**        | `predict(type = "response" | "class")`                                                                |
| Poisson regression                      | `family = "poisson"`                                                                 | count `y`              | **supported**        | offsets supported (`offset =`, `newoffset =`)                                                          |
| Piecewise exponential survival model    | `family = "poisson"` + interval factor + `offset(log(exposure))`                     | event count / interval | **supported (via formulation)** | Cox-like survival modelling expressed as a Poisson regression on long-format data                      |
| Native Cox regression                   | `family = "cox"` + `Surv(time, status) ~ ...`                                        | survival object        | not yet implemented  | `glmnet` supports Cox directly; fbrglm needs family-specific predict handling                          |
| Gamma regression                        | `family = Gamma(link = "log")`                                                       | positive continuous    | future / under investigation | requires glmnet family-object plumbing in fbrglm                                                       |
| Negative binomial regression            | `family = MASS::negative.binomial(theta = ...)`                                      | overdispersed counts   | future / under investigation | fixed-`theta` may be possible via family object; `theta` estimation (à la `MASS::glm.nb`) is separate |

The remaining sections walk through the four **supported** rows in
short reproducible examples. The unsupported rows are shown only as
`eval = FALSE` placeholders at the end.

## Linear regression

```{r linear}
set.seed(101)
n <- 150
dat_lm <- data.frame(
    y  = 0.5 + 0.8 * rnorm(n) - 0.3 * rnorm(n) + rnorm(n, sd = 0.5),
    x1 = rnorm(n),
    x2 = rnorm(n)
)
fit_lm <- fbrglm(y ~ x1 + x2, data = dat_lm,
                 family = "gaussian",
                 lambda = "fix", lambda_value = 0.05)
coef(fit_lm)
head(predict(fit_lm, type = "response"))
```

## Logistic regression

```{r logistic}
set.seed(102)
n <- 200
dat_logit <- data.frame(
    y  = rbinom(n, 1, 0.5),
    x1 = rnorm(n), x2 = rnorm(n)
)
fit_logit <- fbrglm(y ~ x1 + x2, data = dat_logit,
                    family = "binomial",
                    lambda = "fix", lambda_value = 0.05)
head(predict(fit_logit, type = "response"))   # probabilities in [0, 1]
head(predict(fit_logit, type = "class"))      # 0 / 1
```

## Poisson regression with an offset

A common count-data setup attaches an *exposure* (time, area,
population) to each row. The Poisson likelihood is written on the
event count with `log(exposure)` as an additive offset on the linear
predictor.

```{r poisson}
set.seed(103)
n <- 200
exposure <- runif(n, min = 0.5, max = 2)
x1 <- rnorm(n); x2 <- rnorm(n)
rate <- exp(-1 + 0.4 * x1 - 0.2 * x2)
dat_pois <- data.frame(
    y  = rpois(n, rate * exposure),
    x1 = x1, x2 = x2,
    exposure = exposure
)

fit_pois <- fbrglm(y ~ x1 + x2, data = dat_pois,
                   family = "poisson",
                   offset = log(dat_pois$exposure),
                   lambda = "fix", lambda_value = 0.05)

new_dat <- dat_pois[1:5, ]
predict(fit_pois,
        newdata   = new_dat,
        newoffset = log(new_dat$exposure),
        type = "response")
```

## Piecewise exponential survival model via Poisson regression

A Cox-like survival model can be fit with a Poisson regression on a
long-format dataset whose rows are *subject* × *time-interval*. The
construction is:

- split each subject's follow-up time into a small number of
  intervals;
- for each interval the subject is at risk in, generate one row with
  the time at risk (`exposure`), an event indicator (`event`), the
  interval label (a factor), and the time-fixed covariates;
- model `event ~ x1 + x2 + interval + offset(log(exposure))` with a
  Poisson likelihood.

The `interval` factor captures the baseline hazard piecewise; the
`offset(log(exposure))` accounts for time at risk. With sufficiently
fine intervals this formulation is closely related to Cox regression
(the *piecewise exponential* model). It is **not** the same as
`glmnet`'s native Cox partial likelihood, which fbrglm does not yet
expose; see the roadmap below.

```{r pwe-data}
set.seed(104)
n_subj <- 120
breaks <- c(0, 1, 2, 3)
n_int  <- length(breaks) - 1

x1 <- rnorm(n_subj)
x2 <- rnorm(n_subj)
rate    <- exp(-1 + 0.4 * x1 - 0.2 * x2)
T_event <- rexp(n_subj, rate)
T_cens  <- 3
T_obs   <- pmin(T_event, T_cens)
delta   <- as.integer(T_event <= T_cens)

rows <- list()
for (i in seq_len(n_subj)) {
    for (j in seq_len(n_int)) {
        lo <- breaks[j]; hi <- breaks[j + 1]
        if (T_obs[i] <= lo) next               # subject already left
        exposure   <- min(T_obs[i], hi) - lo
        if (exposure <= 0) next
        event_here <- as.integer(delta[i] == 1L &
                                 T_obs[i] >  lo &
                                 T_obs[i] <= hi)
        rows[[length(rows) + 1]] <- data.frame(
            id       = i,
            interval = j,
            exposure = exposure,
            event    = event_here,
            x1       = x1[i],
            x2       = x2[i]
        )
    }
}
long_dat <- do.call(rbind, rows)
long_dat$interval <- factor(long_dat$interval, levels = seq_len(n_int))
str(long_dat)
```

```{r pwe-fit}
fit_pwe <- fbrglm(event ~ x1 + x2 + interval,
                  data   = long_dat,
                  family = "poisson",
                  offset = log(long_dat$exposure),
                  lambda = "fix", lambda_value = 0.05)
coef(fit_pwe)
```

To predict the hazard contribution for a small set of new rows, pass
the matching `newoffset`:

```{r pwe-predict}
new_long <- long_dat[1:6, ]
predict(fit_pwe,
        newdata   = new_long,
        newoffset = log(new_long$exposure),
        type = "response")
```

The `interval` factor is one-hot encoded by `model.matrix()`; if a
prediction-time row happens to omit one of the training intervals, the
corresponding one-hot column is zero-padded by `predict.fbrglm`, as
described in the *Getting started* vignette.

## Roadmap: native Cox, Gamma, negative binomial

These three are **not implemented** in fbrglm yet. The code below
shows the intended API only and does not execute.

### Native Cox

`glmnet` supports `family = "cox"` directly, but fbrglm's
formula/predict pipeline currently has no Cox-specific path
(`Surv(time, status)` on the LHS, partial-likelihood-aware predict
types). Native Cox is a planned milestone.

```{r native-cox, eval = FALSE}
# Reserved but not implemented:
# fbrglm(Surv(time, status) ~ x1 + x2, data = surv_dat,
#        family = "cox", lambda = "cv_min")
```

### Gamma regression

Gamma support depends on glmnet family-object plumbing through
fbrglm's argument validation and predict path; this is currently
under investigation.

```{r gamma, eval = FALSE}
# Under investigation; requires glmnet family-object support in fbrglm:
# fbrglm(y ~ x1 + x2, data = dat,
#        family = Gamma(link = "log"),
#        lambda = "fix", lambda_value = 0.05)
```

### Negative binomial

A fixed-`theta` negative binomial fit may be achievable via a family
object once family-object support lands; `theta` estimation in the
style of `MASS::glm.nb` is a separate problem and is not on the
immediate roadmap.

```{r negbin, eval = FALSE}
# Under investigation; fixed-theta only:
# fbrglm(y ~ x1 + x2, data = dat,
#        family = MASS::negative.binomial(theta = 2),
#        lambda = "fix", lambda_value = 0.05)
```

## Limitations

- The reproducible examples in this vignette are all `gaussian` /
  `binomial` / `poisson`-based; the piecewise exponential survival
  example is also a Poisson formulation, **not** native Cox.
- `infer = "none"` only — no classical standard errors, *z* values,
  *p* values, or confidence intervals are produced. Honest
  post-selection inference (`"split"`, `"selective"`) is planned but
  not implemented.
- Native Cox (`family = "cox"`), Gamma, multinomial, and negative
  binomial families are not yet supported; see the package `TODO.md`.
