---
title: "fbrglm: Safe Formula-Based Workflows for Regularized Generalized Linear Models in R"
author:
  - name: "Koki Tsuyuzaki"
    email: "koki.tsuyuzaki@gmail.com"
    affiliation: chibaDSC
    correspondingauthor: true
    footnote: 1
  - name: "Kentaro Sakamaki"
    email: "kentaro.sakamaki@gmail.com"
    affiliation: juntendo
  - name: "Hiromu Nishiuchi"
    email: "hiromunishiuchi@gmail.com"
    affiliation: chibaDSC
address:
  - code: chibaDSC
    organization: "Data Science Core, Chiba University"
    addressline: "1-33 Yayoi-cho, Inage-ku"
    city: Chiba
    postcode: "263-8522"
    country: Japan
  - code: juntendo
    organization: "Faculty of Health Data Science, Juntendo University"
    city: Tokyo
    country: Japan
footnote:
  - code: 1
    text: "Also affiliated with the Omics AI Research Team, Advanced General Intelligence for Science Program (AGIS), RIKEN TRIP Headquarters, Wako, Japan."
abstract: |
  `glmnet` is the de facto standard for regularized generalized linear
  models in R, but its matrix-shaped API hands factor encoding,
  contrasts, rank handling, and prediction-time design-matrix
  reconstruction back to the caller — producing silent predict-time
  crashes on train / test factor drift, and zero coefficients that
  conflate penalty shrinkage with non-identifiability. `fbrglm` is a
  formula-based wrapper that pairs `glmnet` with a `glm()`-compatible
  workflow: frozen `terms` / `xlevels` / `contrasts` on the fit
  object, NA coefficients for rank-deficient columns, and an inference
  policy that suppresses classical SE / z / p / CI under
  regularization. Predictions are bit-identical to raw `glmnet` across
  all eight supported families.
keywords:
  - regularized generalized linear models
  - glmnet
  - formula interface
  - reproducibility
  - R
journal: "SoftwareX"
date: "`r Sys.Date()`"
linenumbers: false
numbersections: true
bibliography: references.bib
biblio-style: elsarticle-num
classoption: "preprint, 3p"
output:
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
preamble: |
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{array}
  \providecommand{\pkg}[1]{\texttt{#1}}
  \providecommand{\proglang}[1]{\textsf{#1}}
  \providecommand{\code}[1]{\texttt{#1}}
  \emergencystretch=3em
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>",
    fig.width  = 6,
    fig.height = 4,
    echo = TRUE
)
```

```{r metadata, echo = FALSE}
metadata_df <- data.frame(
  Nr = paste0("C", 1:8),
  Description = c(
    "Current code version",
    "Permanent link to code / repository used for this code version",
    "Legal code license",
    "Code versioning system used",
    "Software code languages, tools and services used",
    "Compilation requirements, operating environments and dependencies",
    "If available, link to developer documentation / manual",
    "Support email for questions"
  ),
  Value = c(
    "v0.1.0",
    "https://github.com/dsc-chiba-u/fbrglm",
    "MIT",
    "git",
    "R",
    "R (>= 4.0); glmnet, stats, graphics, generics, tibble",
    "https://CRAN.R-project.org/package=fbrglm",
    "koki.tsuyuzaki@gmail.com"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(metadata_df, format = "latex", booktabs = TRUE,
             col.names = c("Nr", "Code metadata description", "Metadata"),
             caption = "Code metadata (mandatory per SoftwareX template).",
             linesep = "") |>
  kableExtra::column_spec(1, width = "0.6cm") |>
  kableExtra::column_spec(2, width = "5.5cm") |>
  kableExtra::column_spec(3, width = "7cm")
```

# Motivation and significance

Regularized generalized linear models (GLMs) — the L1, L2 and
elastic-net penalized fits provided by `glmnet` [@friedman2010glmnet;
@simon2011glmnetcox; @tay2023glmnetall] — are a workhorse of applied
statistics in R. `glmnet`'s numerical core is mature, but its user
interface is deliberately minimal: the caller supplies a
pre-constructed numeric design matrix `x` and a response `y`, and every
step that `stats::glm()` normally handles on the user's behalf —
formula parsing, factor encoding, contrast bookkeeping, complete-case
filtering, rank checking, and prediction-time design-matrix
reconstruction — is pushed to the caller. In practice this yields
three recurring hazards.

**Predict-time drift.** When the training design matrix is built by
`model.matrix(~ ., data = train)` and the test design matrix by
`model.matrix(~ ., data = test)` independently, a factor level that
appears only in `train` (or only in `test`) silently produces
different numbers of columns, and `predict.glmnet` fails with a
low-level "number of variables in newx must be N" error. In hosted
scoring pipelines this failure mode is common and hard to debug
because the mismatch is between two independently-constructed design
matrices rather than between the model and its input.

**Rank deficiency.** `stats::glm()` reports rank-deficient columns
with `NA` coefficients; `glmnet` accepts collinear columns silently.
A "zero" coefficient in a raw `glmnet` fit can therefore mean either
"shrunk to zero by the penalty" or "not identifiable in the design",
and the two are impossible to distinguish from the fit object alone.

**No inference contract.** Post-selection standard errors for
regularized fits are an unresolved statistical problem
[@tibshirani2016selective], yet callers routinely bolt classical
SE / z / p / CI onto `glmnet` coefficients using the raw penalized
Hessian. Existing wrappers give no explicit signal that these
quantities are unreliable.

`fbrglm` addresses all three by wrapping `glmnet` behind a
`glm()`-compatible formula interface that freezes the training-time
model recipe onto the fit object, restores `glm()`'s rank-deficient
NA convention, and prints an explicit inference policy in place of
untrustworthy classical statistics. The underlying `glmnet` and
`cv.glmnet` calls are unchanged. Interface / workflow packages are
an established category of scientific-software contribution — the
`tidyomics` ecosystem [@hutchison2024tidyomics], for example, was
recognized in *Nature Methods* not for introducing new estimators
but for imposing a unified interface across pre-existing tools.
`fbrglm` occupies the analogous role for `glmnet`.

# Software description

## Software architecture

`fbrglm` exposes two user-facing entry points — `fbrglm()` for a fixed
grid of `lambda` values and its `lambda = "cv_min"` / `"cv_1se"`
shortcuts that delegate to `cv.glmnet` — and a family of `glm()`-style
S3 methods that operate on the returned `regularized_glm` object.

`fbrglm()` parses the formula against the training data, extracts the
`terms` object, `xlevels` (factor levels seen at fit time), and
`contrasts`, builds the numeric design matrix, and hands the pair
`(x, y)` to `glmnet::glmnet` (or `glmnet::cv.glmnet`) verbatim. It
then wraps the returned `glmnet` (or `cv.glmnet`) object together
with the frozen model recipe, a rank check performed by QR-pivot on
the column-centered design matrix (`tol = 1e-7`, matching the
convention used inside `stats::glm.fit`), and a small `nobs_info`
bookkeeping record (`n_total` / `n_dropped_missing` / `n_used`) into a
single `regularized_glm` object.

## Software functionalities

Table 1 lists the S3 methods that operate on a `regularized_glm`.
Each method has direct `glm()` semantics.

```{r methods-table, echo = FALSE}
methods_df <- data.frame(
  Method    = c("print", "summary", "coef", "predict", "nobs",
                "plot", "tidy", "glance",
                "as_glmnet", "as_cv_glmnet"),
  Purpose   = c(
    "compact display of family, lambda, and nobs_info",
    "coefficient table with NA for rank-deficient columns and an explicit inference-policy notice",
    "coefficient vector at the selected lambda; NA preserved for dropped columns",
    "reconstructs the design matrix from stored terms / xlevels / contrasts; supports on_new_levels = 'error' (default) or 'na'",
    "returns n_used from nobs_info",
    "diagnostic residual plots (default), or the underlying glmnet coefficient path / cv.glmnet curve",
    "broom-compatible tidy tibble of coefficients",
    "broom-compatible one-row model-level tibble",
    "accessor returning the underlying glmnet object",
    "accessor returning the underlying cv.glmnet object"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(methods_df, format = "latex", booktabs = TRUE,
             caption = "S3 methods defined on the regularized\\_glm class.",
             linesep = "") |>
  kableExtra::column_spec(1, width = "2.5cm") |>
  kableExtra::column_spec(2, width = "10cm")
```

Supported model families are the full `glmnet` surface — `gaussian`,
`binomial`, `poisson`, `cox`, `multinomial`, `mgaussian` — plus
arbitrary `stats::family` objects such as `stats::Gamma` and
`MASS::negative.binomial` (with a fixed dispersion `theta`). Native
Cox is invoked with a `Surv()` response and `family = "cox"`; the
vignette also walks through the piecewise exponential Poisson
formulation as an alternative.

The prediction path deserves brief comment because it is the point at
which most `glmnet` callers get injured. Given a fitted
`regularized_glm` object and a `newdata` data frame, `predict()`
reconstructs the design matrix by calling `model.matrix` against the
**stored** `terms` and `xlevels` — not against the levels present in
`newdata`. If `newdata` contains a factor level that was not seen at
fit time, `predict()` raises a clear error by default; passing
`on_new_levels = "na"` returns `NA` predictions for the affected rows
instead, which is the appropriate behavior in production scoring
pipelines where scoring must not halt on a single unseen level.

# Illustrative examples

## Binomial regression on `birthwt` with factor drift and rank deficiency

We use `MASS::birthwt` [@MASS], the canonical low-birth-weight cohort
(Springfield, MA, 1986; 189 obstetric records with `low`, an indicator
for birth weight below 2500 g, as the response). It exercises every
safety feature `fbrglm` adds on top of `glmnet`: an ethnicity factor
whose levels vary between train and test slices, a "cleaned" derived
column that turns out to be a linear combination of an existing
factor (a classic data-engineering accident that surfaces as rank
deficiency), and the `glm()`-style coefficient / summary contract on
the fit object.

```{r birthwt-setup, include = FALSE}
## Silently run the same setup shown below so `fit` / `train` / `test`
## exist for the diagnostic-plot chunk.
suppressPackageStartupMessages({library(fbrglm); library(MASS)})
data("birthwt", package = "MASS")
birthwt$race  <- factor(birthwt$race,  1:3, c("white","black","other"))
birthwt$smoke <- factor(birthwt$smoke, 0:1, c("no","yes"))
birthwt$ftv   <- factor(birthwt$ftv)
birthwt$any_visit <- as.integer(birthwt$ftv != "0")
set.seed(93); idx <- sample.int(nrow(birthwt), 150)
train <- birthwt[idx, ]; test <- birthwt[-idx, ]
suppressMessages({
  fit <- fbrglm(low ~ age + lwt + race + smoke + ptl + ftv + any_visit,
                data = train, family = "binomial",
                lambda = "fix", lambda_value = 0.05)
})
```

```{r birthwt, eval = FALSE}
library("fbrglm")
library("MASS")

data("birthwt", package = "MASS")
birthwt$race  <- factor(birthwt$race,  1:3, c("white", "black", "other"))
birthwt$smoke <- factor(birthwt$smoke, 0:1, c("no",    "yes"))
birthwt$ftv   <- factor(birthwt$ftv)

## Data-engineering accident: a downstream ETL job added a "cleaned"
## indicator for any prenatal visit. `any_visit` is a linear function
## of the `ftv` dummies + intercept, so the design matrix is
## rank-deficient once both live in the formula.
birthwt$any_visit <- as.integer(birthwt$ftv != "0")

set.seed(93)
idx   <- sample.int(nrow(birthwt), 150)
train <- birthwt[idx, ]
test  <- birthwt[-idx, ]

fit <- fbrglm(
    low ~ age + lwt + race + smoke + ptl + ftv + any_visit,
    data         = train,
    family       = "binomial",
    lambda       = "fix",
    lambda_value = 0.05
)
#> fbrglm: dropped 1 linearly dependent column(s) (rank 11 / 12):
#>         any_visit. Their coefficients will be reported as NA,
#>         as in stats::glm().

summary(fit)
#> Observations: total = 150, dropped (missing) = 0, used = 150
#>
#> Coefficients: (1 not defined because of singularities: any_visit)
#>                Estimate
#>   (Intercept)   -0.3423
#>   age           -0.0027
#>   lwt           -0.0042
#>   raceblack      0.0000
#>   raceother      0.0000
#>   smokeyes       0.5156
#>   ptl            0.2526
#>   ftv1..ftv6     0.0000     (all shrunk to zero at this lambda)
#>   any_visit           NA
#>
#> Non-zero predictors (4): age, lwt, smokeyes, ptl
#> Note: no standard errors / z / p under infer = "none".
```

Two things to notice. First, `any_visit` shows `NA` in the coefficient
table — `fbrglm` detected the rank deficiency at fit time (QR-pivot
on the column-centered design matrix) and preserved `glm()`'s
convention, so a numeric zero elsewhere in the table unambiguously
means "shrunk by the penalty" rather than "not identifiable". Second,
`summary()` prints only point estimates and an inference-policy notice
in place of SE / z / p / CI; the classical Wald quantities are not
valid post-penalization [@tibshirani2016selective] and the package
refuses to display them by default.

The same `glm()`-shaped API extends to visual diagnostics. Calling
`plot(fit)` returns the three-panel deviance-residual view familiar
from `stats::glm` (Figure 1); raw `glmnet` does not expose
residuals in the `glm()` sense, so this view is a genuine `fbrglm`
add-on rather than a pass-through.

```{r birthwt-diag, echo = TRUE, fig.width = 7.2, fig.height = 1.7, out.width = "100%", fig.pos = "!h", fig.cap = "Standard \\texttt{glm}-style diagnostic panel produced by \\texttt{plot(fit)} on the \\texttt{birthwt} fit. Deviance residuals are computed from the family's \\texttt{dev.resids} at the selected \\texttt{lambda}; the two-arc shape is the binomial GLM signature."}
op <- graphics::par(mfrow = c(1, 3), mar = c(3.5, 3.5, 2, 0.5), mgp = c(2, 0.7, 0))
plot(fit)
graphics::par(op)
```

The split also stresses the predict path: the 39-row test batch
happens to contain 19 `white`, 0 `black`, and 20 `other` observations,
so `model.matrix(~ ., test)` would build a design matrix with fewer
columns than `model.matrix(~ ., train)` — the exact conditions under
which a hand-built `glmnet` call raises "number of variables in newx
must be N". `predict(fit, newdata = test)` succeeds, because the
design matrix is reconstructed against the **stored** `xlevels`
rather than the levels present in `test`; a truly unseen level would
raise a checked error by default, or return `NA` under
`on_new_levels = "na"`.

## Cox regression on `lung` with missing-data reporting

The same call pattern works for native Cox regression. Using
`survival::lung` [@survival2000], several covariates carry natural
missingness that `fbrglm`'s complete-case bookkeeping reports without
having to reason about it externally:

```{r lung-cox, eval = FALSE}
library("survival")
data("lung", package = "survival")
lung$ph.ecog <- factor(lung$ph.ecog)

cox_fit <- fbrglm(
    Surv(time, status) ~ age + sex + ph.ecog + meal.cal + wt.loss,
    data         = lung,
    family       = "cox",
    lambda       = "fix",
    lambda_value = 0.02
)
#> fbrglm: dropped 58 row(s) with missing values (170 / 228 kept).

cox_fit$nobs_info
#> n_total = 228 ;  n_dropped_missing = 58 ;  n_used = 170
```

Any row with an `NA` in the response or in one of the predictors is
dropped, and the exact tally lands on the fit object — the same
`nobs_info` bookkeeping is available across every supported family.
The `vignette("fbrglm-families")` bundled with the package walks the
remaining families end-to-end using the same formula-first pattern.

# Impact

## Improvement to existing regularized-GLM analyses

`fbrglm` is directly applicable to any project that uses `glmnet` for
regularized GLMs — from small ad-hoc analyses to production scoring
pipelines. Four categories of failure that are common in raw-`glmnet`
code disappear at the API level:

- **Predict-time robustness.** Factor-level mismatch between training
  and scoring data becomes an explicit, checked error — or a
  controlled `NA` under `on_new_levels = "na"` — rather than a
  low-level `glmnet` crash. In the companion benchmark
  [@fbrglm-experiments] this class of failure caused 3 of 5 tested
  workflows to fail on a narrowed test factor; `fbrglm` and the
  hardhat-backed `parsnip` workflow are the two paths that survive.
- **Interpretability of "zero".** Rank-deficient columns show as
  `NA` coefficients (`glm()` convention), so a numeric zero in the
  coefficient table unambiguously means "shrunk by the penalty"
  rather than "not identifiable in the design".
- **Inference contract.** The regularization contract is surfaced in
  every `summary()` call, discouraging accidental use of unreliable
  classical Wald statistics for penalized fits and pointing users to
  post-selection inference [@tibshirani2016selective].
- **Interoperability.** `broom` [@broom] methods (`tidy`, `glance`) are
  provided directly, and two accessors return the underlying `glmnet`
  and `cv.glmnet` objects unchanged, so downstream tooling that expects
  a raw `glmnet` handle still works.

## Research workflows enabled

The safety layer changes what regularized GLMs can be trusted with,
not what they compute. Two workflow patterns that were previously
labor-intensive become routine: **hand-off between exploration and
production scoring**, where a serialized fit is applied to arriving
batches without re-materializing the training design matrix; and
**family-wide analyses**, where a study comparing, e.g., binomial,
negative binomial, and Gamma fits on the same cohort no longer needs
a family-specific wrapper per fit — the eight supported families all
share the `regularized_glm` return type. The companion benchmark
[@fbrglm-experiments] uses this uniformity for an
$8 \times 5 = 40$-cell parity sweep against raw `glmnet`,
`glmnetUtils` [@glmnetUtils], and a `parsnip` [@parsnip] /
`workflows` [@workflows] pipeline; every `fbrglm` cell is
bit-identical to raw `glmnet`.

## Changes in daily practice

For a working analyst, the diff against a raw-`glmnet` script is
concentrated in what used to be pure bookkeeping: `model.matrix()`
calls, their re-materialization at predict time, manual `xlevels`
maintenance, and per-family dispatch. All of these move into the
package, leaving the analysis script to say `formula + data + family`
and read back a `regularized_glm` object that behaves like a `glm`
fit for `print`, `summary`, `coef`, `predict`, `nobs`, `plot`, and
`broom` [@broom]. For tidymodels teams, the two `glmnet` accessors
keep `fbrglm` drop-in with existing downstream tooling.

## Adoption and reproducibility

`fbrglm` v0.0.1 was released to CRAN on 2026-07-09 and updated to
v0.1.0 on 2026-07-14; the package has recorded on the order of 200
CRAN downloads across the two releases in its first weeks. Continuous
integration on GitHub runs `R CMD check` on four platforms per push,
and a companion Docker image is republished to Docker Hub and GHCR
on every merge. The runtime overhead of the safety layer is modest:
in a small binomial runtime grid, `fbrglm` sits within roughly
2–3$\times$ the raw-`glmnet` cost — comparable to `glmnetUtils` and
consistently lighter than the `parsnip` / `workflows` path in the
tested setting; these numbers are illustrative rather than a general
benchmark, but they establish that the safety layer is not a numerical
compromise. The release used in the companion study is archived at
[@fbrglm]; the corresponding replication material is at
[@fbrglm-experiments]. The MIT license permits both academic and
commercial use; no commercial deployments are known to the authors at
the time of writing.

# Conclusions

`fbrglm` narrows the gap between `stats::glm()`'s strict
modeling-workflow contract and the numerical machinery of `glmnet`.
By freezing the training-time model recipe on the fit object,
restoring `glm()`'s rank-deficient NA convention, and printing an
explicit inference policy, the package removes a well-known class of
correctness and reproducibility failures that raw `glmnet` invites
but does not itself catch. The underlying estimator is unchanged;
what the user gets is the safety guarantees, error semantics, and
`glm()`-shaped API that make regularized GLMs safe to hand to a
production scoring pipeline.

# References {-}
