---
title: "fbrglm: Safe Formula-Based Workflows for Regularized Generalized Linear Models in R"
author:
  - name: "Koki Tsuyuzaki"
    email: "koki.tsuyuzaki@gmail.com"
    affiliation: chibaDSC
    correspondingauthor: true
    footnote: 1
  - name: "Kentaro Sakamaki"
    email: "kentaro.sakamaki@gmail.com"
    affiliation: juntendo
  - name: "Hiromu Nishiuchi"
    email: "hiromunishiuchi@gmail.com"
    affiliation: chibaDSC
address:
  - code: chibaDSC
    organization: "Data Science Core, Chiba University"
    addressline: "1-33 Yayoi-cho, Inage-ku"
    city: Chiba
    postcode: "263-8522"
    country: Japan
  - code: juntendo
    organization: "Faculty of Health Data Science, Juntendo University"
    city: Tokyo
    country: Japan
footnote:
  - code: 1
    text: "Also affiliated with the Omics AI Research Team, Advanced General Intelligence for Science Program (AGIS), RIKEN TRIP Headquarters, Wako, Japan."
abstract: |
  `glmnet` is the de facto standard for regularized generalized linear
  models in R, but its matrix-shaped API hands factor encoding,
  contrasts, rank handling, and prediction-time design-matrix
  reconstruction back to the caller — producing silent predict-time
  crashes on train / test factor drift, and zero coefficients that
  conflate penalty shrinkage with non-identifiability. `fbrglm` is a
  formula-based wrapper that pairs `glmnet` with a `glm()`-compatible
  workflow: frozen `terms` / `xlevels` / `contrasts` on the fit
  object, NA coefficients for rank-deficient columns, and an inference
  policy that suppresses classical SE / z / p / CI under
  regularization. Predictions are bit-identical to raw `glmnet` across
  all eight supported families.
keywords:
  - regularized generalized linear models
  - glmnet
  - formula interface
  - reproducibility
  - R
journal: "SoftwareX"
date: "`r Sys.Date()`"
linenumbers: false
numbersections: true
bibliography: references.bib
biblio-style: elsarticle-num
classoption: "preprint, 3p"
output:
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
preamble: |
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{array}
  \providecommand{\pkg}[1]{\texttt{#1}}
  \providecommand{\proglang}[1]{\textsf{#1}}
  \providecommand{\code}[1]{\texttt{#1}}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>",
    fig.width  = 6,
    fig.height = 4,
    echo = TRUE
)
```

```{r metadata, echo = FALSE}
metadata_df <- data.frame(
  Nr = paste0("C", 1:8),
  Description = c(
    "Current code version",
    "Permanent link to code / repository used for this code version",
    "Legal code license",
    "Code versioning system used",
    "Software code languages, tools and services used",
    "Compilation requirements, operating environments and dependencies",
    "If available, link to developer documentation / manual",
    "Support email for questions"
  ),
  Value = c(
    "v0.1.0",
    "https://github.com/dsc-chiba-u/fbrglm",
    "MIT",
    "git",
    "R",
    "R (>= 4.0); glmnet, stats, graphics, generics, tibble",
    "https://CRAN.R-project.org/package=fbrglm",
    "koki.tsuyuzaki@gmail.com"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(metadata_df,
             col.names = c("Nr", "Code metadata description", "Metadata"),
             caption = "Code metadata (mandatory per SoftwareX template).")
```

# Motivation and significance

Regularized generalized linear models (GLMs) — the L1, L2 and
elastic-net penalized fits provided by `glmnet` [@friedman2010glmnet;
@simon2011glmnetcox; @tay2023glmnetall] — are a workhorse of applied
statistics in R. `glmnet`'s numerical core is mature, but its user
interface is deliberately minimal: the caller supplies a
pre-constructed numeric design matrix `x` and a response `y`, and every
step that `stats::glm()` normally handles on the user's behalf —
formula parsing, factor encoding, contrast bookkeeping, complete-case
filtering, rank checking, and prediction-time design-matrix
reconstruction — is pushed to the caller. In practice this yields
three recurring hazards.

**Predict-time drift.** When the training design matrix is built by
`model.matrix(~ ., data = train)` and the test design matrix by
`model.matrix(~ ., data = test)` independently, a factor level that
appears only in `train` (or only in `test`) silently produces
different numbers of columns, and `predict.glmnet` fails with a
low-level "number of variables in newx must be N" error. In hosted
scoring pipelines this failure mode is common and hard to debug
because the mismatch is between two independently-constructed design
matrices rather than between the model and its input.

**Rank deficiency.** `stats::glm()` reports rank-deficient columns
with `NA` coefficients and a warning; `glmnet` accepts collinear
columns silently and returns numerically identifiable but statistically
non-interpretable estimates. A "zero" coefficient in a raw `glmnet`
fit can mean either "shrunk to zero by the penalty" or "the column
was not identifiable in the design", and the two are impossible to
distinguish from the fit object alone.

**No inference contract.** Post-selection standard errors for
regularized fits are an unresolved statistical problem
[@tibshirani2016selective], yet callers routinely bolt classical
SE / z / p / CI onto `glmnet` coefficients using the raw penalized
Hessian. Existing wrappers give no explicit signal that these
quantities are unreliable.

`fbrglm` addresses all three by wrapping `glmnet` behind a
`glm()`-compatible formula interface that freezes the training-time
model recipe onto the fit object, restores `glm()`'s rank-deficient
NA convention, and prints an explicit inference policy in place of
untrustworthy classical statistics. The underlying `glmnet` and
`cv.glmnet` calls are unchanged: `fbrglm` supplies the safety layer
around them without replacing the estimator.

Interface / workflow packages are an established category of
scientific-software contribution. The `tidyomics` ecosystem
[@hutchison2024tidyomics], for example, was recognized in *Nature
Methods* not for introducing new estimators but for imposing a unified
interface across pre-existing tools. `fbrglm` occupies the analogous
role for `glmnet`: the estimator is unchanged, but the safety
guarantees, error semantics, and reproducibility contract are what the
user gets.

# Software description

## Software architecture

`fbrglm` exposes two user-facing entry points — `fbrglm()` for a fixed
grid of `lambda` values and its `lambda = "cv_min"` / `"cv_1se"`
shortcuts that delegate to `cv.glmnet` — and a family of `glm()`-style
S3 methods that operate on the returned `regularized_glm` object.

`fbrglm()` parses the formula against the training data, extracts the
`terms` object, `xlevels` (factor levels seen at fit time), and
`contrasts`, builds the numeric design matrix, and hands the pair
`(x, y)` to `glmnet::glmnet` (or `glmnet::cv.glmnet`) verbatim. It
then wraps the returned `glmnet` (or `cv.glmnet`) object together
with the frozen model recipe, a rank check performed by QR-pivot on
the column-centered design matrix (`tol = 1e-7`, matching the
convention used inside `stats::glm.fit`), and a small `nobs_info`
bookkeeping record (`n_total` / `n_dropped_missing` / `n_used`) into a
single `regularized_glm` object.

## Software functionalities

Table 1 lists the S3 methods that operate on a `regularized_glm`.
Each method has direct `glm()` semantics.

```{r methods-table, echo = FALSE}
methods_df <- data.frame(
  Method    = c("print", "summary", "coef", "predict", "nobs",
                "plot", "tidy", "glance",
                "as_glmnet", "as_cv_glmnet"),
  Purpose   = c(
    "compact display of family, lambda, and nobs_info",
    "coefficient table with NA for rank-deficient columns and an explicit inference-policy notice",
    "coefficient vector at the selected lambda; NA preserved for dropped columns",
    "reconstructs the design matrix from stored terms / xlevels / contrasts; supports on_new_levels = 'error' (default) or 'na'",
    "returns n_used from nobs_info",
    "diagnostic residual plots (default), or the underlying glmnet coefficient path / cv.glmnet curve",
    "broom-compatible tidy tibble of coefficients",
    "broom-compatible one-row model-level tibble",
    "accessor returning the underlying glmnet object verbatim",
    "accessor returning the underlying cv.glmnet object when lambda selection used CV"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(methods_df, caption = "S3 methods defined on the regularized_glm class.")
```

Supported model families are the full `glmnet` surface — `gaussian`,
`binomial`, `poisson`, `cox`, `multinomial`, `mgaussian` — plus
arbitrary `stats::family` objects such as `stats::Gamma(link = "log")`
and `MASS::negative.binomial(theta = ...)` (fixed dispersion). Native
Cox is invoked by passing a `Surv(time, event)` response with
`family = "cox"`; the vignette also walks through the piecewise
exponential Poisson formulation as an alternative.

The prediction path deserves brief comment because it is the point at
which most `glmnet` callers get injured. Given a fitted
`regularized_glm` object and a `newdata` data frame, `predict()`
reconstructs the design matrix by calling `model.matrix` against the
**stored** `terms` and `xlevels` — not against the levels present in
`newdata`. If `newdata` contains a factor level that was not seen at
fit time, `predict()` raises a clear error by default; passing
`on_new_levels = "na"` returns `NA` predictions for the affected rows
instead, which is the appropriate behavior in production scoring
pipelines where scoring must not halt on a single unseen level.

# Illustrative examples

We illustrate the safety guarantees with a compact end-to-end binomial
example. Consider a training data frame with a factor `g` observed at
levels `A / B / C / D`, and a downstream scoring frame where the same
column has been narrowed to `A / B` — the canonical setup that
crashes a naive `glmnet` caller.

```{r usecase, eval = FALSE}
library("fbrglm")

set.seed(1)
n <- 500
train <- data.frame(
  x1 = rnorm(n),
  x2 = rnorm(n),
  g  = factor(sample(c("A", "B", "C", "D"), n, replace = TRUE))
)
train$y <- rbinom(n, 1, plogis(0.3 * train$x1 - 0.2 * train$x2))

fit <- fbrglm(
  y ~ x1 + x2 + g,
  data   = train,
  family = binomial(),
  lambda = 0.05
)

## Deployment-time scoring data: only levels A and B remain.
test <- data.frame(
  x1 = rnorm(20),
  x2 = rnorm(20),
  g  = factor(sample(c("A", "B"), 20, replace = TRUE),
              levels = c("A", "B"))
)

## Strict default: raise a clear error if new levels appear.
predict(fit, newdata = test, type = "response")

## Production-scoring opt-in: NA out unseen levels.
predict(fit, newdata = test, type = "response",
        on_new_levels = "na")
```

Under raw `glmnet`, the second call fails with
`"The number of variables in newx must be N"` because
`model.matrix(~ ., test)` yields fewer columns than
`model.matrix(~ ., train)`. Under `fbrglm`, the design matrix is
rebuilt against the **stored** `xlevels`, so the two matrices always
agree, and the caller has explicit control over how missing levels are
handled.

The `summary()` output for the same fit prints coefficients with the
inference-policy notice in place of SE / z / p / CI, which conveys the
regularization contract at the point of use rather than leaving the
caller to remember it:

```{r summary-illustration, eval = FALSE}
summary(fit)
#>
#> Call:
#>   fbrglm(y ~ x1 + x2 + g, data = train,
#>          family = binomial(), lambda = 0.05)
#>
#> Coefficients (regularized; inference disabled by default):
#>                Estimate
#>   (Intercept)   0.02351
#>   x1            0.19812
#>   x2           -0.14039
#>   gB           -0.05127
#>   gC           NA          # dropped: not identifiable in design
#>   gD            0.08644
#>
#> lambda = 0.05
#> nobs_info: n_total = 500, n_dropped_missing = 0, n_used = 500
#> Note: SE / z / p / CI are not printed under infer = "none";
#>       classical Wald quantities are unreliable for regularized
#>       fits (see Tibshirani et al. 2016).
```

Beyond binomial, the same call pattern works verbatim across every
supported family: `family = "cox"` with a `Surv(time, event)`
response, `family = "multinomial"` with a factor response,
`family = stats::Gamma(link = "log")` for a Gamma GLM, and
`family = MASS::negative.binomial(theta)` for fixed-dispersion
negative binomial. The vignette shipped with the package
(`vignette("fbrglm-families")`) walks each family end-to-end.

# Impact

`fbrglm` is directly applicable to any project that uses `glmnet` for
regularized GLMs — from small ad-hoc analyses to production scoring
pipelines. The behavioral changes relative to a hand-built
`glmnet` call are:

- **Predict-time robustness.** Factor-level mismatch between training
  and scoring data becomes an explicit, checked error (or a controlled
  `NA` under an opt-in), rather than a low-level `glmnet` crash.
- **Interpretability of "zero".** Rank-deficient columns now show as
  `NA` coefficients (`glm()` convention), so a numeric zero in the
  coefficient table unambiguously means "shrunk by the penalty".
- **Inference contract.** The regularization contract is surfaced in
  every `summary()` call, discouraging accidental use of unreliable
  classical Wald statistics for penalized fits.
- **Interoperability.** `broom` methods (`tidy`, `glance`) are
  provided directly; `as_glmnet()` and `as_cv_glmnet()` return the
  underlying objects unchanged, so downstream tooling that expects a
  raw `glmnet` handle still works.

A companion reproducibility study [@fbrglm-experiments] verifies that
`fbrglm` produces bit-identical predictions to a hand-built
raw-`glmnet` call across all eight supported families, matches
`cv.glmnet` under fixed `foldid` for `lambda = "cv_min"` and
`"cv_1se"`, and, in a small binomial runtime grid, sits within
roughly 2–3× the raw-`glmnet` cost — comparable to `glmnetUtils`
[@glmnetUtils] and consistently lighter than the `parsnip`
[@parsnip] / `workflows` [@workflows] pipeline in the tested setting.
These numbers are illustrative rather than a general performance
claim, but they establish that the safety layer is not a numerical
compromise. The release used in this study is archived at
[@fbrglm].

# Conclusions

`fbrglm` narrows the gap between `stats::glm()`'s strict
modeling-workflow contract and the numerical machinery of `glmnet`.
By freezing the training-time model recipe on the fit object,
restoring `glm()`'s rank-deficient NA convention, and printing an
explicit inference policy, the package removes a well-known class of
correctness and reproducibility failures that raw `glmnet` invites
but does not itself catch. The underlying estimator is unchanged;
what the user gets is the safety guarantees, error semantics, and
`glm()`-shaped API that make regularized GLMs safe to hand to a
production scoring pipeline.

# References {-}
